---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# `citcdf` <a><img src='man/figures/logo.svg' align="right" height="139" /></a>

[![CRAN status](https://www.r-pkg.org/badges/version/citcdf)](https://CRAN.R-project.org/package=citcdf)
[![R-CMD-check](https://github.com/sistm/citcdf/workflows/R-CMD-check/badge.svg)](https://github.com/sistm/citcdf/actions)

## Overview

`citcdf` is a package to perform conditional independence testing using empirical conditional cumulative distribution function estimations.

The package has two main entry points: `cit_multi()` for gene-wise conditional
independence testing across many outcomes, and `cit_gsa()` for gene-set analysis.
Both accept `test = "asymptotic"` (large samples) or `test = "permutation"`
(small samples). The single-outcome workhorses `cit_asymp()` and `cit_perm()`,
the CCDF estimator `ccdf()`, and the plotting functions `plot_compare_ccdf()`,
`plot.cit_multi()` and `plot.cit_gsa()` are also exported.

The approach implemented in this package is detailed in the following article:

> Gauthier M, Agniel D, Thiébaut R & Hejblum BP (2020). Distribution-free complex hypothesis testing for single-cell RNA-seq differential expression analysis, *BioRxiv*  [doi:10.1101/2021.05.21.445165](https://doi.org/10.1101/2021.05.21.445165)



## Installation


**`citcdf` is available from [GitHub](https://github.com/sistm/citcdf):**
```{r GitHub-install, eval=FALSE}
#install.packages("devtools")
remotes::install_github("sistm/citcdf")
```


## Example

Here is a basic example which shows how to use `citcdf` with simple generated data.

```{r example, eval = FALSE}
## Data Generation
n <- 100
X <- data.frame(X1 = as.factor(rbinom(n = n, size = 1, prob = 0.5)))
Y <- replicate(10, (X$X1 == 1) * rnorm(n) + (X$X1 == 0) * rnorm(n, mean = 0.5))
```

```{r estimation, eval = FALSE}
# Hypothesis testing
res_asymp <- cit_multi(M = data.frame(Y = Y), X = X,
                       test = "asymptotic", parallel = FALSE) # asymptotic test
res_asymp$pvals

res_perm <- cit_multi(M = data.frame(Y = Y), X = X,
                      test = "permutation", adaptive = TRUE,
                      parallel = FALSE) # adaptive permutation test
res_perm$pvals
```



 -- Marine Gauthier, Denis Agniel, Kalidou Ba, Rodolphe Thiébaut & Boris Hejblum
 
*hex illustration by Jérôme Dubois.*


